Amortization of Sub-programs in Simulators

### 2019-09-07 paper/learning/notes.tex
add tom notes

\subsection{3rd Baseline}

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=\textwidth]{tom_derivation.jpg}
  \caption{Tom's derivation proving that the $\mathcal{L}_2$ norm in this problem set-up enables us to learn a proposal that exactly
  matches the true distribution.  }
  \label{fig:tomfig}
\end{figure}
\section{ Questions for Tom}

@@

\section{Additional TODOs}

>>>>

% \subsection{3rd Baseline}

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=\textwidth]{tom_derivation.jpg}
  \caption{Tom's derivation proving that the $\mathcal{L}_2$ norm in this problem set-up enables us to learn a proposal that exactly
  matches the true distribution.  }
  \label{fig:tomfig}
\end{figure}

\subsection{Are the samples in Method 3 Unbiased?}
The short answer is that we seem to be able to do something that will give better per-iteration performance than vanilla LMH, 
but that we have an extra term to calculate that means we don't also get the speedups in how much each iteration costs that we might have hoped for.

The expected number of cycles through the rejection sampler $\mathbb{E}[N_a | \lambda]$,
is equal to 1 over the acceptance ratio of the sampler $Z(\lambda)$ as per the proof we did earlier. \textbf{Bradley add proof}
This means that we can use the $N_a$ to learn a direct approximator for $\frac{1}{Z(\lambda)}$ for method 3.

Let $P^{in}(z)$ denote the density of the rejection sampler with the conditioning statement removed 
(i.e. the density on the outputs if we just accepted all samples).  
Let $a(z)=1$ denote the event of acceptance. 
Let $p(z)$ denote the density of the outputs from running the rejection sampler.

We now have the following results
\begin{align}
   p(z ; \lambda) =& P^{in}(z ; \lambda) \frac{\mathbb{I}(a(z ; \lambda)=1)}{Z(\lambda)} \\
   \mathbb{E}[P^{in}(z ; \lambda) \mathbb{I}(a(z ; \lambda)=1) N_a | z] =& P^{in}(z ; \lambda) \mathbb{I}(a(z ; \lambda)=1) \mathbb{E}[N_a | z, \lambda]\\
=& P^{in}(z ; \lambda) \mathbb{I}(a(z ; \lambda)=1) \mathbb{E}[N_a | \lambda] \\ %(see explanation below) 
=& p(z ; \lambda) = P^{in}(z ; \lambda) \frac{\mathbb{I}(a(z ; \lambda)=1)}{Z(\lambda)}
\end{align}

The key here, that I was worrying about before, is the second step
that is does $\mathbb{E}[N_a | z, \lambda] = \mathbb{E}[N_a | \lambda]$?
The answer is that it does if we think about things from the perspective as $z$
as the output from the rejection sampler. 
The exact value of this has no impact on $N_a$, only the fact that it was accepted.
But, by definition, we know that $z$ has been accepted, so  $\mathbb{E}[N_a | z, \lambda]$
is the same for all possible $z$ and thus equal to  $\mathbb{E}[N_a | \lambda]$.

This argument is perhaps easiest to see in reverse: we can think of first sampling $N_a \sim P(N_a)$
to figure out how long we sample for before acceptance, then we can sample an output using $z \sim p(z ; \lambda)$, which clearly does not depend on $N_a$ so the two must be independent.
In other words, though the full set of samples that are generated during running $x_{1:N_a}$ is clearly not independent from $N_a$, the final sample we return $z$ definitely is.
Note that this does not cause any complications with the fact that we never actually evaluate $P^{in}(z)$, but its preimage $P^{in}(x_{N_a})$
instead, because the result further implies that $x_{N_a}$ and $N_a$ are independent, even though $x_{1:N_a}$ and $N_a$ are not.

In general, it is best to think about pseudo-marginal schemes (i.e. MCMC chains based on unbiased density estimates) as estimating the density by drawing a
random number seed $\epsilon \sim p(\epsilon)$, then producing a deterministic density for a given parameter value and this random number seed. 
We can thus think of the density estimate as being constructed through the process

\begin{align}
  \epsilon &\sim p(\epsilon)\text{(draw a seed)} \\
  \hat{p}(z) &= p(z;\epsilon)\text{(i.e. the estimate is a deterministic function of $z$ and epsilon)}
\end{align}

If we now do an MCMC step, our implied proposal on $\{\epsilon,z\}$ space becomes $q(\epsilon',z'|\epsilon,z) = p(\epsilon')q(z'|z)$ for some $q(z'|z)$ that we choose.
Using our density estimates now gives the pseudo-marginal MH acceptance ratio
\begin{equation}
  P_a = \frac{ \hat{p}(z') q(z | z') }{ \hat{p}(z') p(z;\epsilon) q(z' | z) }
\end{equation}
and from the previous results we have
\begin{align}
&=  \frac{ p(z';\epsilon') q(z | z') }{ p(z;\epsilon) q(z' | z) } \\
&= \frac{ p(z';\epsilon') q(\epsilon,z|\epsilon',z')/p(\epsilon)  }{ p(z;\epsilon) q(\epsilon',z'|\epsilon,z)/ p(\epsilon')  } \\
&=  \frac{ p(z';\epsilon') p(\epsilon') q(\epsilon,z|\epsilon',z') }{ p(z;\epsilon) p(\epsilon)  q(\epsilon',z'|\epsilon,z) }
\end{align}
which is a valid MH sampler on the unnormalized density $p(\epsilon) p(z;\epsilon)$.
If our $\hat{p}(z)$ are unbiased, this implies that $\mathbb{E}[ \hat{p}(z) | z] = \int p(epsilon) p(z;\epsilon)  d\epsilon = p(z)$,
hence $p(z)$ is the marginal distribution of $ p(\epsilon) p(z;\epsilon)$, so drawing samples from  $p(\epsilon) p(z;\epsilon)$  produces the desired distribution on z and the pseudo-marginal MCMC sampler is valid.

We can now view our case in the general pseudo-marginal setting above.  
Effectively we have $\epsilon \sim p(\epsilon)$ (draw a random number seed for the local rejection sampler)
$N_a = N_a(\epsilon; \lambda)$, (i.e. $N_a$ is a deterministic function of $\epsilon$ and $\lambda$)
$\hat{p}(z) = p_in(z ; \lambda) \mathbb{I}(a(z ; \lambda)=1) N_a(\epsilon; \lambda)$

This is perfectly fine to run a pseudo-marginal scheme on, but the problem is that if $\lambda$ changes, so does $N_a$.  
Thus even though we could theoretically use our trick about $N_a$ to run things in a pseudo marginal manner with, e.g., 
LMH updates where we keep the $\epsilon$ and $z$ fixed for rejection samplers downstream from the one we are updating, 
the change in $\lambda$ means not only do we need to recalculate $P^{in}(z ; \lambda) \mathbb{I}(a(z ; \lambda)$ 
as we would normally do but which only requires a single pass through the rejection sampler, we also have to recalculate $N_a (\epsilon; \lambda)$,
which requires us to pass through the rejection sampler multiple times, such that our updates remain extremely expensive. 
Alternatively, we could also just redraw $N_a$ for all downstream rejection samplers (with a bit of care that I haven't not fully gone through),
which effectively equates to redrawing the epsilons and is thus simply a different MH update.
However, again this means we need to repass through the rejection sampling loops.

Interestingly, there is still potentially some gain to be had here as we have circumvented the need to redraw the rejection sampler outputs (i.e. $z$) of future draws in the way that we would need to for vanilla LMH.
As such, this would give us "method 0b" and it should be more efficient than vanilla LMH as even though the updates remain expensive, 
it does avoid the curse of dimensionality issues we get from having to redraw all the future $z$'s from the prior each time we update an earlier one: 
we only have to recalculate the number of cycles to acceptance and then use these in the acceptance ratio (and next iteration if they are accepted).

The per-iteration cost of this will be similar to an iteration of training a regressor to predict $N_a$, but much higher than running inference once we have learned such a regressor.  
There is also potential to combine the two methods: running pseudo-marginal LMH will generate $(\lambda,N_a)$ sample pairs that can be used for the training so you could initially run 
pseudo-marginal LMH and at some point switch when you're confident you have learned a good enough regressor switch to relying on that instead.

@@

\section{ Questions for Tom}
\section{Additional TODOs}
